---
title: "Family Movie Ratings Analysis 3a: Back To The Future"
author: "Randy Howk"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
  html_document:
    toc: true
    toc_float: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.width = 10, fig.height = 6)

# Source setup utilities
source("setup.R")

# Test database connection and setup
test_db_connection()

# Connect to database
con <- connect_to_movie_db()
```

# Introduction

This analysis examines movie ratings from our family for 6 recent popular movies. The participants are:

- **Randy** (56 Father)
- **Elle** (42, Mother) 
- **Ariah** (9, Daughter)
- **Alexandra** (6, Daughter)
- **Don** (82, Grandfather)

The movies rated include both focus films (Barbie and KPop Demon Hunters) plus 4 additional recent popular movies suitable for a wide age range. In addition, we will attempt to use Global Baseline Estimates about the shows and the viewers.

# Data Loading

## Load Raw Data from Database

```{r load-data}
# Get all ratings with participant and movie details
ratings_data <- get_ratings_with_names(con)

# Display the dataset
knitr::kable(ratings_data, 
             caption = "Complete Movie Ratings Dataset",
             col.names = c("Name", "Age", "Relationship", "Movie", "Year", "Genre", "Rating", "Notes"))
```

## Summary Statistics

```{r summary-stats}
# Get movie averages
movie_averages <- get_movie_averages(con)

knitr::kable(movie_averages,
             caption = "Movie Rating Averages",
             col.names = c("Movie Title", "Number of Ratings", "Average Rating", "Min Rating", "Max Rating"))

# Overall statistics
cat("Dataset Summary:\n")
cat("Total ratings collected:", nrow(ratings_data), "\n")
cat("Number of movies:", length(unique(ratings_data$movie_title)), "\n") 
cat("Number of participants:", length(unique(ratings_data$person_name)), "\n")
cat("Average rating across all movies:", round(mean(ratings_data$rating), 2), "\n")
```

# Data Analysis

## Rating Distribution by Movie

```{r rating-distribution}
# Create bar chart of average ratings by movie
movie_avg_plot <- movie_averages %>%
  ggplot(aes(x = reorder(title, avg_rating), y = avg_rating, fill = title)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = avg_rating), hjust = -0.1) +
  coord_flip() +
  labs(title = "Average Movie Ratings",
       subtitle = "Family ratings on scale of 1-5",
       x = "Movie",
       y = "Average Rating") +
  theme_minimal() +
  scale_y_continuous(limits = c(0, 5.5))

print(movie_avg_plot)
```

## Ratings by Age Group

```{r age-analysis}
# Create age groups for analysis
ratings_with_age_groups <- ratings_data %>%
  mutate(age_group = case_when(
    age <= 10 ~ "Children (6-9)",
    age <= 50 ~ "Adults (42-45)", 
    age > 50 ~ "Senior (82+)"
  ))

# Average rating by age group
age_group_summary <- ratings_with_age_groups %>%
  group_by(age_group) %>%
  summarise(
    avg_rating = round(mean(rating), 2),
    count = n(),
    .groups = 'drop'
  )

knitr::kable(age_group_summary,
             caption = "Average Ratings by Age Group")

# Plot ratings by age group and movie
age_movie_plot <- ratings_with_age_groups %>%
  ggplot(aes(x = movie_title, y = rating, fill = age_group)) +
  geom_boxplot(alpha = 0.7) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Movie Ratings by Age Group",
       x = "Movie",
       y = "Rating (1-5)",
       fill = "Age Group")

print(age_movie_plot)
```

## Individual Rating Patterns

```{r individual-patterns}
# Heatmap of individual ratings
individual_plot <- ratings_data %>%
  ggplot(aes(x = movie_title, y = person_name, fill = rating)) +
  geom_tile(color = "white") +
  geom_text(aes(label = rating), color = "white", fontface = "bold") +
  scale_fill_gradient2(low = "red", mid = "yellow", high = "green", 
                      midpoint = 3, limits = c(1,5)) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Individual Movie Ratings Heatmap",
       x = "Movie",
       y = "Family Member",
       fill = "Rating")

print(individual_plot)
```

## Focus Movies Analysis

```{r focus-movies}
# Focus on the two focus movies
focus_movies <- ratings_data %>%
  filter(movie_title %in% c("Barbie", "KPop Demon Hunters"))

knitr::kable(focus_movies %>% select(person_name, age, movie_title, rating, notes),
             caption = "Ratings for focus Movies (Barbie & KPop Demon Hunters)")

# Compare the focus movies
focus_comparison <- focus_movies %>%
  group_by(movie_title) %>%
  summarise(
    avg_rating = round(mean(rating), 2),
    ratings_range = paste(min(rating), "to", max(rating)),
    .groups = 'drop'
  )

knitr::kable(focus_comparison,
             caption = "focus Movies Comparison")
```

# Key Findings

```{r findings}
# Find highest and lowest rated movies
best_movie <- movie_averages$title[which.max(movie_averages$avg_rating)]
worst_movie <- movie_averages$title[which.min(movie_averages$avg_rating)]

# Find most generous and harsh raters
person_averages <- ratings_data %>%
  group_by(person_name, age) %>%
  summarise(avg_rating = round(mean(rating), 2), .groups = 'drop') %>%
  arrange(desc(avg_rating))

most_generous <- person_averages$person_name[1]
most_critical <- person_averages$person_name[nrow(person_averages)]

cat("Key Findings:\n")
cat("• Highest rated movie:", best_movie, 
    "(", max(movie_averages$avg_rating), "average )\n")
cat("• Lowest rated movie:", worst_movie, 
    "(", min(movie_averages$avg_rating), "average )\n")
cat("• Most generous rater:", most_generous, 
    "(", max(person_averages$avg_rating), "average )\n") 
cat("• Most critical rater:", most_critical, 
    "(", min(person_averages$avg_rating), "average )\n")
```

# Global Baseline Estimate Recommendation System

## Overview

The Global Baseline Estimate is a non-personalized recommendation algorithm that predicts ratings based on:
- The overall average rating across all movies and users
- How each user tends to rate compared to the average (user bias)
- How each movie tends to be rated compared to the average (movie bias)

## Implementation

```{r global-baseline-setup}
# Function to implement Global Baseline Estimate
calculate_global_baseline <- function(ratings_data) {
  # Calculate overall mean rating
  mean_all_ratings <- mean(ratings_data$rating, na.rm = TRUE)
  
  # Calculate user statistics and biases
  user_stats <- ratings_data %>%
    group_by(person_name) %>%
    summarise(
      user_avg = mean(rating, na.rm = TRUE),
      n_ratings = n(),
      .groups = 'drop'
    ) %>%
    mutate(user_bias = user_avg - mean_all_ratings)
  
  # Calculate movie statistics and biases
  movie_stats <- ratings_data %>%
    group_by(movie_title) %>%
    summarise(
      movie_avg = mean(rating, na.rm = TRUE),
      n_ratings = n(),
      .groups = 'drop'
    ) %>%
    mutate(movie_bias = movie_avg - mean_all_ratings)
  
  return(list(
    mean_all = mean_all_ratings,
    user_stats = user_stats,
    movie_stats = movie_stats
  ))
}

# Function to predict a rating
predict_rating <- function(user, movie, baseline_model) {
  user_bias <- baseline_model$user_stats %>%
    filter(person_name == user) %>%
    pull(user_bias)
  
  movie_bias <- baseline_model$movie_stats %>%
    filter(movie_title == movie) %>%
    pull(movie_bias)
  
  # Handle missing values
  if (length(user_bias) == 0) user_bias <- 0
  if (length(movie_bias) == 0) movie_bias <- 0
  
  # Calculate prediction
  predicted <- baseline_model$mean_all + user_bias + movie_bias
  
  # Constrain to 1-5 scale
  return(max(1, min(5, predicted)))
}
```

# Model Training

```{r global-model-training}
# Train the model
baseline_model <- calculate_global_baseline(ratings_data)

# Display overall statistics
cat("Global Mean Rating:", round(baseline_model$mean_all, 2), "\n\n")

# Display user biases
knitr::kable(baseline_model$user_stats %>% 
               arrange(desc(user_bias)) %>%
               mutate(across(where(is.numeric), ~round(., 2))),
             caption = "User Rating Patterns and Biases")

# Display movie biases
knitr::kable(baseline_model$movie_stats %>% 
               arrange(desc(movie_bias)) %>%
               mutate(across(where(is.numeric), ~round(., 2))),
             caption = "Movie Rating Patterns and Biases")
```

# Generating Predictions

```{r generating-predictions}
# Create all possible user-movie combinations
all_combinations <- expand.grid(
  person_name = unique(ratings_data$person_name),
  movie_title = unique(ratings_data$movie_title),
  stringsAsFactors = FALSE
)

# Add predictions
predictions <- all_combinations %>%
  mutate(
    predicted_rating = mapply(predict_rating, person_name, movie_title, 
                             MoreArgs = list(baseline_model = baseline_model))
  ) %>%
  left_join(ratings_data %>% select(person_name, movie_title, actual_rating = rating),
            by = c("person_name", "movie_title"))

# Show sample predictions
sample_predictions <- predictions %>%
  filter(!is.na(actual_rating)) %>%
  sample_n(min(10, n())) %>%
  mutate(
    error = actual_rating - predicted_rating,
    abs_error = abs(error)
  ) %>%
  arrange(person_name, movie_title)

knitr::kable(sample_predictions %>%
               mutate(across(where(is.numeric), ~round(., 2))),
             caption = "Sample Predictions vs Actual Ratings")
```

# Model Performance

```{r model-performance}
# Calculate performance metrics on known ratings
known_ratings <- predictions %>%
  filter(!is.na(actual_rating))

mae <- mean(abs(known_ratings$actual_rating - known_ratings$predicted_rating))
rmse <- sqrt(mean((known_ratings$actual_rating - known_ratings$predicted_rating)^2))

cat("Model Performance Metrics:\n")
cat("- Mean Absolute Error (MAE):", round(mae, 3), "\n")
cat("- Root Mean Square Error (RMSE):", round(rmse, 3), "\n")
cat("- Number of predictions evaluated:", nrow(known_ratings), "\n")

# Visualize prediction accuracy
accuracy_plot <- known_ratings %>%
  ggplot(aes(x = actual_rating, y = predicted_rating)) +
  geom_point(alpha = 0.6, size = 3) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "red") +
  scale_x_continuous(breaks = 1:5, limits = c(0.5, 5.5)) +
  scale_y_continuous(breaks = 1:5, limits = c(0.5, 5.5)) +
  labs(title = "Prediction Accuracy: Actual vs Predicted Ratings",
       x = "Actual Rating",
       y = "Predicted Rating") +
  theme_minimal() +
  coord_equal()

print(accuracy_plot)
```

# Recommendations

```{r recoommendations}
# Generate recommendations for each user
recommendations <- predictions %>%
  filter(is.na(actual_rating)) %>%
  group_by(person_name) %>%
  arrange(desc(predicted_rating)) %>%
  slice_head(n = 2) %>%
  select(person_name, movie_title, predicted_rating) %>%
  mutate(predicted_rating = round(predicted_rating, 2))

knitr::kable(recommendations,
             caption = "Top 2 Movie Recommendations for Each Family Member")

# Focus on specific example (similar to Excel spreadsheet)
# Let's predict what Don would rate Barbie
example_user <- "Don"
example_movie <- "Barbie"

if (example_user %in% ratings_data$person_name && 
    example_movie %in% ratings_data$movie_title) {
  
  user_bias <- baseline_model$user_stats %>%
    filter(person_name == example_user) %>%
    pull(user_bias)
  
  movie_bias <- baseline_model$movie_stats %>%
    filter(movie_title == example_movie) %>%
    pull(movie_bias)
  
  predicted <- predict_rating(example_user, example_movie, baseline_model)
  
  cat("\n### Example Prediction Detail\n")
  cat("Predicting", example_user, "'s rating for", example_movie, "\n")
  cat("- Global mean rating:", round(baseline_model$mean_all, 2), "\n")
  cat("- ", example_user, "'s bias (tendency to rate higher/lower):", round(user_bias, 2), "\n")
  cat("- ", example_movie, "'s bias (tends to be rated higher/lower):", round(movie_bias, 2), "\n")
  cat("- Predicted rating:", round(baseline_model$mean_all, 2), "+", 
      round(user_bias, 2), "+", round(movie_bias, 2), "=", round(predicted, 2), "\n")
}
```

# Recommendation-insights

```{r recommendation-insights}
# Analyze which movies are most recommended
most_recommended <- predictions %>%
  filter(is.na(actual_rating)) %>%
  group_by(movie_title) %>%
  summarise(
    avg_predicted = mean(predicted_rating),
    times_recommended = sum(predicted_rating >= 4),
    .groups = 'drop'
  ) %>%
  arrange(desc(avg_predicted))

knitr::kable(most_recommended %>%
               mutate(avg_predicted = round(avg_predicted, 2)),
             caption = "Movies Most Likely to be Enjoyed by Family Members Who Haven't Seen Them")
```

# Technical Details

```{r tech-details}
# Show database structure
tables <- dbListTables(con)
cat("DDEV Database tables:", paste(tables, collapse = ", "), "\n")

# Show connection info
cat("\nDatabase connection details:\n")
cat("Host: 127.0.0.1\n")
cat("Port: 3306\n")
cat("Database: db\n")
cat("User: db\n")
```

```{r cleanup, include=FALSE}
# Clean up connection
dbDisconnect(con)
```

# Methodology Notes

- **Data Collection**: Family members rated movies on a scale of 1-5
- **Storage**: Ratings stored in MariaDB database via DDEV
- **Analysis**: Loaded into R dataframes using RMariaDB package
- **Reproducibility**: Complete DDEV setup ensures consistent environment across platforms

---
*Analysis completed on `r Sys.Date()` using R, MariaDB, and DDEV*
