---
title: "Assignment10a-Sentiment Analysis with Tidy Data"
author: "Randy Howk"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: united
    highlight: tango
---

# Introduction

This document demonstrates sentiment analysis using tidy data principles. The primary example code is based on Chapter 2 of "Text Mining with R: A Tidy Approach" by Julia Silge and David Robinson【N:cmhhademz01cwoe0fwgxngx3b】.

# Setup

First, we load the necessary libraries and get the sentiment lexicons.

```{r setup, include=FALSE}
library(dplyr)
library(tidytext)
library(ggplot2)
library(stringr)
library(janeaustenr)
library(tidyr)
library(tm)
library(sentimentr)
```
# Example Code
Here we work with Jane Austen's novels and perform sentiment analysis using the bing lexicon.

```{r bing-lexicon, include=FALSE}
tidy_books <- austen_books() %>%
  group_by(book) %>%
  mutate(
    linenumber = row_number(),
    chapter = cumsum(str_detect(text, regex("^chapter [\\divxlc]", ignore_case = TRUE)))
  ) %>%
  ungroup() %>%
  unnest_tokens(word, text)

nrc_joy <- get_sentiments("nrc") %>% 
  filter(sentiment == "joy")

joy_words <- tidy_books %>%
  filter(book == "Emma") %>%
  inner_join(nrc_joy) %>%
  count(word, sort = TRUE)

joy_words
```
# Extending the Example
## Different Corpus
Next, we work with a different corpus. Replacing Gödel, Escher, Bach extracted separately from PDF.
```{r godel-escher-bach-processing}
# Assumed text extraction from PDF—process similar to plain text if already extracted.
godel_escher_bach <- readLines("GEB_extracted.txt")

# Check and remove any invalid UTF-8 characters
godel_escher_bach <- iconv(godel_escher_bach, from = "UTF-8", to = "UTF-8", sub = "")

# Create a tibble
godel_escher_bach_df <- tibble(line = 1:length(godel_escher_bach), text = godel_escher_bach)

# Unnest tokens
tidy_godel_escher_bach <- godel_escher_bach_df %>%
  unnest_tokens(word, text)

# Print a sample of the processed data
print(head(tidy_godel_escher_bach))


geb_sentiment <- tidy_godel_escher_bach %>%
  inner_join(get_sentiments("bing")) %>%
  count(word, sentiment, sort = TRUE)

geb_sentiment
```
# Additional Sentiment Lexicon
## Incorporate an additional sentiment lexicon from another R package, such as sentimentr.

```{r sentence-extraction}
library(sentimentr)

# Load and clean the text data
godel_escher_bach <- readLines("GEB_extracted.txt", encoding = "UTF-8")
godel_escher_bach <- iconv(godel_escher_bach, from = "UTF-8", to = "UTF-8", sub = "")
godel_escher_bach <- trimws(godel_escher_bach)
godel_escher_bach <- godel_escher_bach[godel_escher_bach != ""] 

# Create tibble/Dataframe from cleaned text
godel_escher_bach_df <- tibble(line = 1:length(godel_escher_bach), text = godel_escher_bach)

# Segment the text into sentences
sentence_list <- get_sentences(godel_escher_bach_df$text)

# Flatten and ensure consistency with sentiment analysis
flattened_sentences <- unlist(sentence_list, recursive = FALSE)

# Perform sentiment analysis
sentences_data <- sentiment_by(flattened_sentences)

# Sample 10 random sentences
sample_ids <- sample(sentences_data$element_id, 10)

# Map and extract the sentences corresponding to the sampled element_ids
sample_sentences <- sentences_data[sentences_data$element_id %in% sample_ids, ]
samples_with_text <- cbind(sample_sentences, text = flattened_sentences[sample_ids])

# Print each sentence individually
cat("## Sample Sentences from Gödel, Escher, Bach\n\n")
for(i in 1:nrow(samples_with_text)) {
  cat(sprintf("**Sentence %d (ID: %d):**\n", i, samples_with_text$element_id[i]))
  cat(samples_with_text$text[i], "\n\n")
}

# Display the statistics table at the end
cat("## Sentiment Analysis Statistics Table\n\n")
knitr::kable(samples_with_text[, c("element_id", "word_count", "sd", "ave_sentiment")], 
             col.names = c("Element ID", "Word Count", "SD", "Average Sentiment"),
             caption = "Sentiment Statistics for Sample Sentences")

```
# Conclusion
This assignment extended the analysis by applying sentiment analysis to a different corpus, GEB, and incorporating an additional sentiment lexicon beyond those discussed in Chapter 2 of the source text.

# References

Hofstadter, D. R. (1979). *Gödel, Escher, Bach: An Eternal Golden Braid*. Basic Books.

Electronic version retrieved from https://welib.org/slow_download/06372a47c23884e660e534a0cc872f9b/0/0 (Accessed: `r format(Sys.Date(), "%B %d, %Y")`)

: Silge, J., & Robinson, D. (2025). Text Mining with R: A Tidy Approach.