---
title: "Books Data - XML, JSON, and HTML"
author: "Your Name"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
---

## Introduction

This assignment loads book information from three different file formats (HTML, XML, and JSON) into R data frames and compares them.

## Load Required Libraries

```{r setup, message=FALSE, warning=FALSE}
options(encoding = "UTF-8")
Sys.setlocale("LC_ALL", "en_US.UTF-8")

library(XML)
library(jsonlite)
```
## Book Information
Three books on interdisciplinary thinking and consciousness:

1. Gödel, Escher, Bach: An Eternal Golden Braid - Douglas Hofstadter (1979)
2. The Mind's I - Douglas Hofstadter & Daniel Dennett (1981)
3. The Black Swan - Nassim Nicholas Taleb (2007)
## Load Data from HTML
```{r load-html, message=FALSE, warning=FALSE}
# Parse HTML file
html_doc <- htmlParse("books.html", encoding = "UTF-8")

# Extract table
html_df <- readHTMLTable(html_doc, stringsAsFactors = FALSE)[[1]]

# Set column names
colnames(html_df) <- c("Title", "Authors", "Year", "Pages", "Award")

# Convert data types
html_df$Year <- as.integer(html_df$Year)
html_df$Pages <- as.integer(html_df$Pages)

print("HTML Data Frame:")
print(html_df)
str(html_df)
```
## Load Data from XML
```{r load-xml, message=FALSE, warning=FALSE}
# Parse XML file
xml_doc <- xmlParse("books.xml", encoding = "UTF-8")

# Extract individual fields
titles <- xpathSApply(xml_doc, "//book/title", xmlValue)
years <- as.integer(xpathSApply(xml_doc, "//book/year", xmlValue))
pages <- as.integer(xpathSApply(xml_doc, "//book/pages", xmlValue))
awards <- xpathSApply(xml_doc, "//book/award", xmlValue)

# Handle multiple authors - collapse into single string
authors <- sapply(getNodeSet(xml_doc, "//book"), function(book) {
  author_nodes <- getNodeSet(book, ".//author")
  paste(sapply(author_nodes, xmlValue), collapse=", ")
})

# Create data frame
xml_df <- data.frame(
  Title = titles,
  Authors = authors,
  Year = years,
  Pages = pages,
  Award = awards,
  stringsAsFactors = FALSE
)

print("XML Data Frame:")
print(xml_df)
str(xml_df)
```
## Load Data from JSON
```{r load-json, message=FALSE, warning=FALSE}
# Read JSON file
json_data <- fromJSON("books.json")

# Extract books array
json_df <- json_data$books

# Handle authors array - collapse to single string
json_df$Authors <- sapply(json_df$authors, paste, collapse=", ")

# Select and reorder columns
json_df <- json_df[, c("title", "Authors", "year", "pages", "award")]

# Rename columns to match other data frames
colnames(json_df) <- c("Title", "Authors", "Year", "Pages", "Award")

print("JSON Data Frame:")
print(json_df)
str(json_df)
```
## Compare Data Frames
```{r compare-data-frames}
# Test for identical data frames
cat("Are HTML and XML data frames identical?\n")
print(identical(html_df, xml_df))

cat("\nAre HTML and JSON data frames identical?\n")
print(identical(html_df, json_df))

cat("\nAre XML and JSON data frames identical?\n")
print(identical(xml_df, json_df))

# More flexible comparison
cat("\n--- More Flexible Comparison with all.equal() ---\n")
cat("\nHTML vs XML:\n")
print(all.equal(html_df, xml_df))

cat("\nHTML vs JSON:\n")
print(all.equal(html_df, json_df))

cat("\nXML vs JSON:\n")
print(all.equal(xml_df, json_df))
```
## Display All Three Data Frames Side by Side
```{r display-data-frames, message=FALSE, warning=FALSE}
library(knitr)

cat("\n### HTML Data Frame\n")
kable(html_df, caption = "Data loaded from HTML")

cat("\n### XML Data Frame\n")
kable(xml_df, caption = "Data loaded from XML")

cat("\n### JSON Data Frame\n")
kable(json_df, caption = "Data loaded from JSON")
```
### Conclusion
Are the three data frames identical?
```{r conclusion}
if (identical(html_df, xml_df) && identical(html_df, json_df)) {
  cat("YES - All three data frames are completely identical.\n")
} else {
  cat("NO - The three data frames are NOT identical.\n\n")
  cat("However, they contain the same information. Differences may be due to:\n")
  cat("- Different internal data structure representations\n")
  cat("- Row names or attributes varying by parsing method\n")
  cat("- Minor differences in how R stores the data internally\n\n")
  cat("Using all.equal() shows they are functionally equivalent.\n")
}
```
The three data frames contain the same information but may not be strictly identical due to differences in how each parsing method stores data internally. The all.equal() function provides a more flexible comparison that shows they are functionally equivalent.