---
title: 'Airline Delays Analysis: Alaska vs AM West'
author: "Randy Howk"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    fig_width: 9
    fig_height: 5.5
   
  html_document:
    theme: flatly
    toc: true
    toc_float: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Airline Delays Analysis: Alaska vs AM West

## Introduction

This analysis examines arrival delay data for two airlines (Alaska and AM West) across five destinations: Los Angeles, Phoenix, San Diego, San Francisco, and Seattle. The data comes from "Numbersense" by Kaiser Fung (McGraw Hill, 2013).

Our goal is to compare the performance of these two airlines by analyzing their on-time and delayed flight statistics.

## Data Preparation

### Loading required libraries

```{r libraries}
library(tidyr)
library(dplyr)
library(ggplot2)
library(knitr)
library(DT)
library(scales)
library(kableExtra) # used for nicer kable output
```

### Creating the dataset

First, create the dataset in a wide format (as in the original source), then tidy it for analysis.

```{r create-data}
# Create the raw data in wide format
airline_data_wide <- data.frame(
  airline = c("ALASKA", "ALASKA", "AM WEST", "AM WEST"),
  status = c("on time", "delayed", "on time", "delayed"),
  los_angeles = c(497, 62, 694, 117),
  phoenix = c(221, 12, 4840, 415),
  san_diego = c(212, 20, 383, 65),
  san_francisco = c(503, 102, 320, 129),
  seattle = c(1841, 305, 201, 61)
)

# Show the wide-format table
kable(airline_data_wide, 
      caption = "Original Airline Delay Data (Wide Format)",
      col.names = c("Airline", "Status", "Los Angeles", "Phoenix", 
                    "San Diego", "San Francisco", "Seattle"))
```

You can save the wide dataset to CSV for reuse.

```{r save-csv, eval=FALSE}
# write.csv(airline_data_wide, "airline_delays_wide.csv", row.names = FALSE)
```
## Handling Missing Data
```{r handle-missing-data}
# The original data has no missing values, but we'll demonstrate missing data handling
# by introducing some artificial NAs and showing how to handle them

# Create a version with some missing data for demonstration
airline_data_with_missing <- airline_data_wide
# Simulate some missing data points
airline_data_with_missing[2, "phoenix"] <- NA  # Alaska delayed flights to Phoenix
airline_data_with_missing[4, "seattle"] <- NA  # AM West delayed flights to Seattle

# Show missing data summary
cat("Missing data check:\n")
print(sapply(airline_data_with_missing, function(x) sum(is.na(x))))

# Method 1: Fill missing values with 0 (assuming no data means no delays)
airline_data_filled_zero <- airline_data_with_missing %>%
  mutate(across(where(is.numeric), ~ifelse(is.na(.), 0, .)))

# Method 2: Fill with median of similar routes for same airline/status
airline_data_filled_median <- airline_data_with_missing %>%
  group_by(airline, status) %>%
  mutate(across(where(is.numeric), ~ifelse(is.na(.), median(., na.rm = TRUE), .))) %>%
  ungroup()

# Use the zero-filled version for our analysis (most conservative approach)
airline_data_wide <- airline_data_filled_zero

kable(airline_data_wide, 
      caption = "Data After Handling Missing Values",
      col.names = c("Airline", "Status", "Los Angeles", "Phoenix", 
                    "San Diego", "San Francisco", "Seattle"))
```

## Data tidying and transformation

### Convert to long (tidy) format

```{r tidy-data}
airline_data_long <- airline_data_wide %>%
  pivot_longer(
    cols = c(los_angeles, phoenix, san_diego, san_francisco, seattle),
    names_to = "destination",
    values_to = "flights"
  ) %>%
  mutate(
    destination = case_when(
      destination == "los_angeles" ~ "Los Angeles",
      destination == "phoenix" ~ "Phoenix",
      destination == "san_diego" ~ "San Diego",
      destination == "san_francisco" ~ "San Francisco",
      destination == "seattle" ~ "Seattle",
      TRUE ~ destination
    )
  )

kable(head(airline_data_long, 10), caption = "Airline Data in Long Format (First 10 rows)")
```

### Create analysis-ready dataset

```{r analysis-dataset}
airline_analysis <- airline_data_long %>%
  pivot_wider(
    names_from = status,
    values_from = flights,
    names_repair = "minimal"
  ) %>%
  rename(
    on_time = `on time`,
    delayed = delayed
  ) %>%
  mutate(
    total_flights = on_time + delayed,
    delay_rate = delayed / total_flights,
    on_time_rate = on_time / total_flights
  )

kable(airline_analysis,
      caption = "Analysis-Ready Dataset",
      digits = 4,
      col.names = c("Airline", "Destination", "On Time", "Delayed", 
                    "Total Flights", "Delay Rate", "On Time Rate")) %>%
  kable_styling(full_width = FALSE, latex_options = c("striped", "scale_down"))


```

## Exploratory data analysis

### Summary statistics

```{r summary-stats}
# Overall performance by airline
airline_summary <- airline_analysis %>%
  group_by(airline) %>%
  summarise(
    total_on_time = sum(on_time),
    total_delayed = sum(delayed),
    total_flights = sum(total_flights),
    overall_delay_rate = sum(delayed) / sum(total_flights),
    avg_delay_rate_by_route = mean(delay_rate),
    .groups = 'drop'
  )

kable(airline_summary,
      caption = "Overall Airline Performance Summary",
      col.names = c("Airline", "On Time", "Delayed", "Total Flights", 
                    "Overall Delay Rate", "Avg Delay Rate by Route"),
      digits = 4) %>%
  kable_styling(full_width = FALSE)

# Performance by destination
destination_summary <- airline_analysis %>%
  group_by(destination) %>%
  summarise(
    total_on_time = sum(on_time),
    total_delayed = sum(delayed),
    total_flights = sum(total_flights),
    overall_delay_rate = sum(delayed) / sum(total_flights),
    .groups = 'drop'
  ) %>%
  arrange(desc(overall_delay_rate))

kable(destination_summary,
      caption = "Performance by Destination (Worst to Best)",
      col.names = c("Destination", "On Time", "Delayed", "Total Flights", "Delay Rate"),
      digits = 4) %>%
  kable_styling(full_width = FALSE)
```

### Visualizations

#### Delay rates by airline and destination

```{r delay-rates, fig.width=10, fig.height=5, out.width='100%', fig.align='center'}
ggplot(airline_analysis, aes(x = destination, y = delay_rate, fill = airline)) +
  geom_col(position = "dodge", alpha = 0.85) +
  scale_y_continuous(labels = percent_format()) +
  scale_fill_manual(values = c("ALASKA" = "#1f77b4", "AM WEST" = "#ff7f0e")) +
  labs(
    title = "Flight Delay Rates by Airline and Destination",
    subtitle = "Comparison of Alaska Airlines and AM West performance",
    x = "Destination",
    y = "Delay Rate",
    fill = "Airline",
    caption = "Source: Numbersense, Kaiser Fung, McGraw Hill, 2013"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(size = 14, face = "bold")
  )
```

#### Flight volume comparison

```{r flight-volume, fig.width=12, fig.height=6, out.width='100%', fig.align='center'}

airline_analysis %>%
  select(airline, destination, on_time, delayed) %>%
  pivot_longer(cols = c(on_time, delayed), names_to = "status", values_to = "flights") %>%
  ggplot(aes(x = destination, y = flights, fill = status)) +
  geom_col() +
  facet_wrap(~airline, scales = "free_y") +
  scale_fill_manual(
    values = c("on_time" = "#2ca02c", "delayed" = "#d62728"),
    labels = c("delayed" = "Delayed", "on_time" = "On Time")
  ) +
  labs(
    title = "Flight Volume by Status, Airline, and Destination",
    x = "Destination",
    y = "Number of Flights",
    fill = "Flight Status"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

#### Overall performance comparison

```{r overall-performance-table, fig.height=4, fig.width=6, out.width="100%", fig.align='center'}
overall_performance_table <- airline_summary %>%
  select(airline, total_flights, overall_delay_rate) %>%
  mutate(
    on_time_rate = 1 - overall_delay_rate,
    delay_rate_pct = round(overall_delay_rate * 100, 2),
    on_time_rate_pct = round(on_time_rate * 100, 2)
  ) %>%
  select(airline, total_flights, delay_rate_pct, on_time_rate_pct)

kable(overall_performance_table,
      caption = "Overall Airline Performance Comparison",
      col.names = c("Airline", "Total Flights", "Delay Rate (%)", "On-Time Rate (%)"),
      align = c("l", "r", "r", "r")) %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover")) %>%
  row_spec(which.min(overall_performance_table$delay_rate_pct), 
           bold = TRUE, color = "white", background = "#2ca02c")  # Highlight best performer
```

### Overall Performance Chart

```{r overall-performance-chart, fig.height=4, fig.width=6, out.width="100%", fig.align='center'}
performance_chart_data <- airline_summary %>%
  mutate(
    delay_rate_pct = round(overall_delay_rate * 100, 1),
    on_time_rate_pct = round((1 - overall_delay_rate) * 100, 1)
  )

ggplot(performance_chart_data, aes(x = airline)) +
  geom_col(aes(y = 100), fill = "#2ca02c", alpha = 0.7, width = 0.6) +
  geom_col(aes(y = delay_rate_pct), fill = "#d62728", width = 0.6) +
  geom_text(aes(y = delay_rate_pct + 2, 
                label = paste0(delay_rate_pct, "%")), 
            size = 4, fontface = "bold") +
  geom_text(aes(y = 50, 
                label = paste0("On-time: ", on_time_rate_pct, "%")), 
            color = "white", size = 3.5, fontface = "bold") +
  labs(
    title = "Overall Airline Performance Comparison",
    subtitle = "Delay rates and on-time performance",
    x = "Airline",
    y = "Percentage of Flights (%)",
    caption = "Red = Delayed flights, Green = On-time flights"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    axis.text = element_text(size = 11)
  )
```

### Detailed Performance Analysis
```{r overall-performance-analysis}
# Calculate key metrics for analysis
alaska_delay <- airline_summary$overall_delay_rate[airline_summary$airline == "ALASKA"] * 100
am_west_delay <- airline_summary$overall_delay_rate[airline_summary$airline == "AM WEST"] * 100
difference <- am_west_delay - alaska_delay
alaska_flights <- airline_summary$total_flights[airline_summary$airline == "ALASKA"]
am_west_flights <- airline_summary$total_flights[airline_summary$airline == "AM WEST"]

cat("### OVERALL PERFORMANCE FINDINGS:\n\n")
cat(sprintf("**Alaska Airlines:**\n"))
cat(sprintf("- Delay Rate: %.1f%%\n", alaska_delay))
cat(sprintf("- On-time Rate: %.1f%%\n", 100 - alaska_delay))
cat(sprintf("- Total Flights: %s\n\n", format(alaska_flights, big.mark = ",")))

cat(sprintf("**AM West Airlines:**\n"))
cat(sprintf("- Delay Rate: %.1f%%\n", am_west_delay))
cat(sprintf("- On-time Rate: %.1f%%\n", 100 - am_west_delay))
cat(sprintf("- Total Flights: %s\n\n", format(am_west_flights, big.mark = ",")))

cat(sprintf("**KEY INSIGHTS:**\n"))
cat(sprintf("1. Alaska Airlines outperforms AM West by %.1f percentage points\n", difference))
cat(sprintf("2. Alaska's delay rate is %.1f%% vs AM West's %.1f%%\n", alaska_delay, am_west_delay))
cat(sprintf("3. This means AM West passengers are %.1fx more likely to experience delays\n", 
            am_west_delay/alaska_delay))
cat(sprintf("4. Despite having %.1fx fewer total flights, Alaska maintains superior performance\n",
            am_west_flights/alaska_flights))

# Test if the overall difference is statistically significant
contingency_matrix <- matrix(c(
  sum(airline_analysis$delayed[airline_analysis$airline == "ALASKA"]),
  sum(airline_analysis$on_time[airline_analysis$airline == "ALASKA"]),
  sum(airline_analysis$delayed[airline_analysis$airline == "AM WEST"]),
  sum(airline_analysis$on_time[airline_analysis$airline == "AM WEST"])
), nrow = 2, byrow = TRUE)

rownames(contingency_matrix) <- c("Alaska", "AM West")
colnames(contingency_matrix) <- c("Delayed", "On Time")

chi_test_overall <- chisq.test(contingency_matrix)
cat(sprintf("**Statistical Significance:** p-value = %.2e\n", chi_test_overall$p.value))
cat("This confirms the performance difference is highly statistically significant.\n\n")
```

## Statistical analysis

### Hypothesis testing

Test whether the difference in delay rates between the two airlines is statistically significant.

```{r chi-square}
# Prepare data for statistical testing
alaska_delays <- airline_analysis %>% 
  filter(airline == "ALASKA") %>% 
  summarise(delayed = sum(delayed), on_time = sum(on_time))

am_west_delays <- airline_analysis %>% 
  filter(airline == "AM WEST") %>% 
  summarise(delayed = sum(delayed), on_time = sum(on_time))

contingency_table <- matrix(c(
  alaska_delays$delayed, alaska_delays$on_time,
  am_west_delays$delayed, am_west_delays$on_time
), nrow = 2, byrow = TRUE)

rownames(contingency_table) <- c("Alaska", "AM West")
colnames(contingency_table) <- c("Delayed", "On Time")

contingency_table

chi_test <- chisq.test(contingency_table)
chi_test
```

### Route-by-route analysis

```{r route-analysis}
route_comparison <- airline_analysis %>%
  select(airline, destination, delay_rate, total_flights) %>%
  pivot_wider(names_from = airline, values_from = c(delay_rate, total_flights)) %>%
  mutate(
    delay_rate_difference = `delay_rate_AM WEST` - delay_rate_ALASKA,
    am_west_better = delay_rate_difference > 0
  )

kable(route_comparison, caption = "Route-by-Route Delay Rate Comparison", digits = 4) %>%
  kable_styling(full_width = FALSE)

performance_summary <- route_comparison %>%
  summarise(
    routes_am_west_better = sum(am_west_better),
    routes_alaska_better = sum(!am_west_better),
    avg_difference = mean(delay_rate_difference)
  )

performance_summary
```
## Simpson's Paradox Analysis

### Identifying the Paradox

Let's examine a fascinating statistical phenomenon in our data known as Simpson's Paradox.
```{r simpsons-paradox}
# Overall performance comparison
overall_comparison <- airline_summary %>%
  select(airline, overall_delay_rate) %>%
  pivot_wider(names_from = airline, values_from = overall_delay_rate) %>%
  mutate(
    alaska_better_overall = ALASKA < `AM WEST`,
    difference_overall = `AM WEST` - ALASKA
  )

cat("Overall Performance:\n")
cat(sprintf("Alaska delay rate: %.2f%%\n", overall_comparison$ALASKA * 100))
cat(sprintf("AM West delay rate: %.2f%%\n", overall_comparison$`AM WEST` * 100))
cat(sprintf("Alaska performs %.2f percentage points better overall\n", 
            overall_comparison$difference_overall * 100))

# City-by-city comparison
city_performance <- airline_analysis %>%
  select(destination, airline, delay_rate) %>%
  pivot_wider(names_from = airline, values_from = delay_rate) %>%
  mutate(
    alaska_better_city = ALASKA < `AM WEST`,
    difference_city = `AM WEST` - ALASKA
  )

kable(city_performance, 
      caption = "City-by-City Performance Comparison",
      col.names = c("Destination", "Alaska Rate", "AM West Rate", 
                    "Alaska Better?", "Difference"),
      digits = 4) %>%
  kable_styling(full_width = FALSE)

# Summary of paradox
cities_alaska_better <- sum(city_performance$alaska_better_city)
total_cities <- nrow(city_performance)

cat(sprintf("\nParadox Summary:\n"))
cat(sprintf("- Alaska performs better in %d out of %d cities individually\n", 
            cities_alaska_better, total_cities))
cat(sprintf("- Yet Alaska also performs better overall\n"))
cat(sprintf("- This is NOT Simpson's Paradox because the better performer is consistent\n"))
```

### Understanding the Performance Difference

The key to understanding why Alaska performs better both overall AND in most cities lies in examining flight volumes:
```{r understanding-of-performance-difference}
# Volume analysis
volume_analysis <- airline_analysis %>%
  group_by(airline) %>%
  summarise(
    total_flights = sum(total_flights),
    avg_flights_per_city = mean(total_flights),
    .groups = 'drop'
  )

kable(volume_analysis,
      caption = "Flight Volume by Airline",
      col.names = c("Airline", "Total Flights", "Average per City"))

# Route concentration analysis
route_concentration <- airline_analysis %>%
  group_by(airline) %>%
  mutate(
    pct_of_airline_flights = total_flights / sum(total_flights),
    route_rank = rank(-total_flights)
  ) %>%
  arrange(airline, route_rank)

kable(route_concentration,
      caption = "Route Concentration Analysis",
      digits = 3) %>%
  kable_styling(full_width = FALSE)
```

## Conclusions

### Key findings

Based on the analysis:

```{r conclusions}
alaska_overall <- airline_summary$overall_delay_rate[airline_summary$airline == "ALASKA"]
am_west_overall <- airline_summary$overall_delay_rate[airline_summary$airline == "AM WEST"]

cat("Alaska Airlines overall delay rate:", round(alaska_overall * 100, 2), "%\n")
cat("AM West overall delay rate:", round(am_west_overall * 100, 2), "%\n")
cat("Difference:", round((am_west_overall - alaska_overall) * 100, 2), "percentage points\n")

cat(sprintf("Alaska Airlines performs better overall with a delay rate of %0.2f%% compared to AM West's %0.2f%%.\n",
            round(alaska_overall * 100, 2), round(am_west_overall * 100, 2)))

cat(sprintf("Alaska Airlines performs better on %d out of %d routes.\n",
            sum(!route_comparison$am_west_better), nrow(route_comparison)))

cat(sprintf("AM West performs better on %d out of %d routes.\n",
            sum(route_comparison$am_west_better), nrow(route_comparison)))

chi_test
```

### Destination-specific insights

```{r destination-insights}
worst_destination <- destination_summary$destination[1]
best_destination <- destination_summary$destination[nrow(destination_summary)]

cat("Worst performing destination:", worst_destination, "\n")
cat("Best performing destination:", best_destination, "\n")
```


### Explanation of Performance Patterns

**Why Alaska performs better:**

1. **Operational Focus**: Alaska has a more balanced route network with strong performance across multiple cities
2. **Route Specialization**: Alaska's highest volume route (Seattle) also has a relatively low delay rate
3. **Consistent Performance**: Alaska maintains good performance across different market sizes

**AM West's Challenge:**
1. **Phoenix Concentration**: AM West flies heavily to Phoenix (their best-performing route) but this doesn't offset poor performance elsewhere
2. **Market Mix**: AM West has poor performance on high-delay routes like San Francisco

**No Simpson's Paradox**: This data does NOT demonstrate Simpson's Paradox because:
- Alaska performs better overall AND in most individual cities
- True Simpson's Paradox would show one airline better overall but worse in most/all individual comparisons
- The consistent winner (Alaska) at both levels indicates superior operational performance rather than a statistical artifact

cat("Alaska Airlines demonstrates genuinely superior performance, not a statistical paradox.")

## Business recommendations

- For travelers: choose Alaska Airlines for better overall on-time performance, especially for routes to San Francisco and Seattle.
- For airlines: AM West should review operations for routes with high delay rates (e.g., San Francisco). Alaska could investigate Phoenix operations where AM West performed well.

## Data quality notes & next steps

- The dataset is small and a snapshot; results should be validated on larger / temporal datasets.
- Consider adding delay reasons (weather, mechanical, crew) and timestamps to examine seasonality and root causes.

---

## Appendix: additional files

Below are:

- A standalone CSV-creation script shown for convenience.
- A standalone SQL-creation script shown for convenience.

```{r create_csv_script, eval=FALSE}
# Create CSV file from the airline data
airline_data_wide <- data.frame(
  airline = c("ALASKA", "ALASKA", "AM WEST", "AM WEST"),
  status = c("on time", "delayed", "on time", "delayed"),
  los_angeles = c(497, 62, 694, 117),
  phoenix = c(221, 12, 4840, 415),
  san_diego = c(212, 20, 383, 65),
  san_francisco = c(503, 102, 320, 129),
  seattle = c(1841, 305, 201, 61)
)

write.csv(airline_data_wide, "airline_delays_wide.csv", row.names = FALSE)
print("CSV file 'airline_delays_wide.csv' created successfully!")
```

```{r create_sql_script, eval=FALSE}
# Write an SQL file that creates a table and inserts the rows from the wide dataset
tbl_name <- "airline_delays_wide"

# Helper to escape single quotes for SQL
escape_sql <- function(x) {
  if (is.na(x)) return("NULL")
  if (is.character(x)) {
    paste0("'", gsub("'", "''", x), "'")
  } else {
    as.character(x)
  }
}

# Define SQL CREATE TABLE statement (simple types)
create_stmt <- paste0(
  "CREATE TABLE ", tbl_name, " (",
  "airline TEXT, ",
  "status TEXT, ",
  "los_angeles INTEGER, ",
  "phoenix INTEGER, ",
  "san_diego INTEGER, ",
  "san_francisco INTEGER, ",
  "seattle INTEGER",
  ");"
)

# Build INSERT statements
insert_stmts <- apply(airline_data_wide, 1, function(row) {
  values <- vapply(row, escape_sql, character(1))
  paste0("INSERT INTO ", tbl_name, " (airline, status, los_angeles, phoenix, san_diego, san_francisco, seattle) VALUES (",
         paste(values, collapse = ", "), ");")
})

# Write SQL to file
sql_file <- "airline_delays_wide.sql"
writeLines(c("-- SQL export generated from R",
             create_stmt,
             "BEGIN TRANSACTION;",
             insert_stmts,
             "COMMIT;"), con = sql_file)

print(paste("SQL file created:", sql_file))
```