---
title: "Analysis of Trump Administration LGBTQ+ Policy Actions"
author: "Randy Howk"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    number_sections: true
    latex_engine: xelatex
    keep_tex: false
geometry: margin=0.75in
header-includes:
  - \usepackage{longtable}
  - \usepackage{booktabs}
  - \usepackage{array}
  - \usepackage{float}
  - \usepackage{caption}
  - \captionsetup[figure]{font=small}
  - \captionsetup[table]{font=small}
  - \usepackage{fvextra}
  - \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,breakanywhere,commandchars=\\\{\}}
---

# Executive Summary

This R Markdown documents a reproducible workflow to convert an untidy Word document of policy actions into a tidy dataset, perform Named Entity Recognition (NER), and conduct exploratory analysis. Database interaction chunks are set to eval=FALSE so the file can knit without a live PostgreSQL connection.

```{r setup, message=FALSE, warning=FALSE}

options(repos = c(CRAN = "https://cran.rstudio.com/"))
library(tidyverse)
library(lubridate)
library(DBI)
library(RPostgres)
library(officer)
library(xml2)
library(tidytext)
library(stringr)
library(ggplot2)
library(scales)
library(knitr)
library(kableExtra)

knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 10, fig.height = 6,tidy.opts = list(width.cutoff = 60), tidy = TRUE)
```

## Step 1: Extract Text from Word Document
Using the officer package to read .docx files directly in R:

```{r extract-docx}
# Adjust the path if your file location differs
docx_path <- "Untidy.docx"

if (!file.exists(docx_path)) {
  stop("Please place Untidy.docx in the working directory or update 'docx_path'.")
}

doc <- read_docx(docx_path)
doc_content <- docx_summary(doc)
text_paragraphs <- doc_content %>% 
  filter(content_type == "paragraph") %>% 
  pull(text) %>% 
  paste(collapse = "\n")

cat("First 500 characters of extracted text:\n")
cat(substr(text_paragraphs, 1, 500), "\n...\n")
```

## Step 2: Tidy and Transform Data

Understanding the "Untidy" Data Structure
The source data from Untidy.docx is untidy for several key reasons that violate Hadley Wickham's tidy data principles:

Unstructured text format: All data is embedded in free-form paragraphs with no clear delimiters or column structure
Date-content concatenation: Dates and descriptive content are merged in single text strings
Inconsistent formatting: No standardized schema or field separation
Non-rectangular structure: The original Word document contains flowing text rather than tabular data
According to tidy data principles, each variable should form a column, each observation should form a row, and each type of observational unit should form a table. Our raw data violates all three principles.

### Split Text into Individual Entries
```{r tidy-transformation-step-1}
# Step 1: Split continuous text into individual entries
# Using dplyr's pipe operator (%>%) for readable data transformation pipeline

raw_entries <- str_split(text_paragraphs, "\n")[[1]] %>%   # Split on line breaks
  tibble(raw_text = .) %>%                                   # Convert to tibble (tidy data frame)
  filter(nchar(trimws(raw_text)) > 0)                       # Remove empty/whitespace-only lines

cat("Total raw entries extracted:", nrow(raw_entries), "\n\n")

# Display sample of raw, untidy data
cat("Example of UNTIDY raw data:\n")
print(head(raw_entries$raw_text, 3))
cat("\n")
```

### Tidying Process: Applying tidyr and dplyr Principles
Now we apply systematic transformations following the tidyr philosophy of reshaping data and dplyr verbs for data manipulation:

```{r tidy-transformation-step-2}
# Step 2: Parse and separate variables into distinct columns
# This is the core "tidying" transformation

policy_data <- raw_entries %>%
  
  # MUTATE: Create new variables by transforming existing ones (dplyr verb)
  mutate(
    
    # Variable 1: Extract the DATE component
    # Pattern: ^\d{2}\.\d{2}\.\d{4} means "start of line (^), two digits, 
    # period, two digits, period, four digits"
    # Example: "09.24.2025" matches this pattern
    date_str = str_extract(raw_text, "^\\d{2}\\.\\d{2}\\.\\d{4}"),
    
    # Variable 2: Extract the CONTENT component
    # Remove the date prefix to isolate the policy description
    # str_remove eliminates the matched pattern from the beginning
    content = str_remove(raw_text, "^\\d{2}\\.\\d{2}\\.\\d{4}\\s*"),
    
    # Variable 3: Convert date STRING to proper DATE data type
    # mdy() from lubridate parses "month.day.year" format
    # This transforms character data into a true date object for temporal analysis
    date = mdy(date_str),
    
    # Variable 4: Create unique identifier for each observation
    # row_number() generates sequential IDs, treating each row as a distinct observation
    entry_id = row_number()
    
  ) %>%
  
  # FILTER: Keep only complete cases with valid dates (dplyr verb)
  # Removes malformed entries that don't match our expected structure
  filter(!is.na(date)) %>%
  
  # SELECT: Choose final variables and reorder columns (dplyr verb)
  # Creates our final tidy structure: one observation per row, one variable per column
  select(entry_id, date, content)

# Report on the tidying results
cat("\n=== TIDYING TRANSFORMATION RESULTS ===\n")
cat("Original raw entries:", nrow(raw_entries), "\n")
cat("Successfully parsed entries:", nrow(policy_data), "\n")
cat("Entries filtered out (no valid date):", nrow(raw_entries) - nrow(policy_data), "\n")
cat("\nDate range captured:", 
    format(min(policy_data$date), "%B %d, %Y"), 
    "to", 
    format(max(policy_data$date), "%B %d, %Y"), "\n")
```
### Demonstrating the Transformation
Let's compare the untidy vs. tidy structure side-by-side:

```{r tidy-transformation-demonstration}
# Show first 3 entries in UNTIDY format
cat("\n--- BEFORE TIDYING (Untidy Format) ---\n")
cat("Single text blob with embedded structure:\n\n")
for(i in 1:3) {
  cat(paste0("[Entry ", i, "] ", substr(raw_entries$raw_text[i], 1, 150), "...\n\n"))
}

cat("\n--- AFTER TIDYING (Tidy Format) ---\n")
cat("Structured table with separated variables:\n\n")

# Display first 3 entries in TIDY format as a formatted table
kable(
  head(policy_data, 3) %>%
    mutate(content_display = substr(content, 1, 100)) %>%
    mutate(content_display = paste0(content_display, "...")) %>%
    select(entry_id, date, content_display),
  caption = "Tidy Data Structure: Each Variable in Its Own Column",
  col.names = c("ID", "Date", "Content (truncated)"),
  format = "latex",
  booktabs = TRUE,
  longtable = TRUE
) %>%
  kable_styling(latex_options = c("striped", "scale_down", "repeat_header"),     font_size = 9) %>%
  column_spec(1, bold = TRUE, width = "5em") %>%
  column_spec(2, width = "10em") %>%
  column_spec(3, width = "40em")
```

### Key Tidying Operations Applied
This transformation leverages multiple tidyr and dplyr principles:

1. Separation (tidyr principle): Date and content separated into distinct columns using str_extract() and str_remove()
2. Type Conversion: Dates transformed from character strings to proper Date objects using lubridate::mdy()
3. Filtering (dplyr verb): Only valid observations with parseable dates retained using filter(!is.na(date))
4. Selection and Ordering (dplyr verb): Essential variables chosen in logical order using select()
5. Unique Identification: Each observation given a unique entry_id using row_number()

```{r tidy-transformation-larger-sample}
# Display larger sample of tidy data with full formatting
kable(
  head(policy_data, 10),
  caption = "First 10 Policy Entries (Tidied and Structured)",
  col.names = c("Entry ID", "Date", "Policy Action Description"),
  format = "latex",
  booktabs = TRUE,
  longtable = TRUE
) %>%
  kable_styling(
    latex_options = c("striped", "scale_down", "repeat_header"),     font_size = 9,
    full_width = FALSE,
    position = "left"
  ) %>%
  column_spec(1, bold = TRUE, border_right = TRUE) %>%
  column_spec(2, bold = TRUE, color = "blue") %>%
  scroll_box(height = "400px")
```

### Data Quality Validation
```{r tidy-transformation-data-quality-validation}
# Verify the tidied data meets quality standards

# Check 1: No missing dates (critical variable)
missing_dates <- sum(is.na(policy_data$date))
cat("Missing dates:", missing_dates, 
    ifelse(missing_dates == 0, "✓ PASS\n", "✗ FAIL\n"))

# Check 2: No missing content
missing_content <- sum(is.na(policy_data$content) | policy_data$content == "")
cat("Missing content:", missing_content, 
    ifelse(missing_content == 0, "✓ PASS\n", "✗ FAIL\n"))

# Check 3: Unique entry IDs
duplicate_ids <- sum(duplicated(policy_data$entry_id))
cat("Duplicate IDs:", duplicate_ids, 
    ifelse(duplicate_ids == 0, "✓ PASS\n", "✗ FAIL\n"))

# Check 4: Date range is logical
date_range_valid <- max(policy_data$date) >= min(policy_data$date)
cat("Date range valid:", 
    ifelse(date_range_valid, "✓ PASS\n", "✗ FAIL\n"))

# Check 5: Data types are correct
correct_types <- is.integer(policy_data$entry_id) && 
                 inherits(policy_data$date, "Date") && 
                 is.character(policy_data$content)
cat("Column types correct:", 
    ifelse(correct_types, "✓ PASS\n", "✗ FAIL\n"))

# Summary statistics
cat("\n--- Summary Statistics ---\n")
cat("Total observations:", nrow(policy_data), "\n")
cat("Variables:", ncol(policy_data), "\n")
cat("Memory size:", format(object.size(policy_data), units = "Kb"), "\n")

# Display structure
cat("\n--- Data Structure ---\n")
str(policy_data)
```
### Why This Tidying Matters for Analysis
The transformation from untidy text to tidy tabular data enables:

1. Temporal analysis: Proper Date objects allow time-series analysis, trend detection, and chronological ordering
2. Text mining: Separated content field enables tokenization, sentiment analysis, and keyword extraction
3. Database integration: Rectangular structure can be directly loaded into SQL databases
4. Statistical modeling: Tidy format is required for most R statistical functions
5. Reproducibility: Structured data can be easily filtered, grouped, and summarized using dplyr verbs

This tidying process converts unstructured narrative text into analysis-ready data, following the core principle that "tidy datasets are all alike, but every messy dataset is messy in its own way."

## Step 3: Connect to PostgreSQL via DDEV
Establishing connection to the local DDEV PostgreSQL instance:

```{r db-connect}
# Connection parameters for DDEV PostgreSQL
con <- dbConnect(
  RPostgres::Postgres(),
  host = "127.0.0.1",           # DDEV exposes postgres on localhost
  port = 55432,                 # DDEV custom port
  user = "ddev",                # Default DDEV user
  password = "ddev",            # Default DDEV password
  dbname = "ddev"               # Default DDEV database
)

# Test connection
cat("Database connection successful!\n")
cat("PostgreSQL version:", dbGetQuery(con, "SELECT version();")$version, "\n")
```

## Step 4: Load Tidied Data into PostgreSQL
Writing the tidied data to the database:

```{r db-write}
# Write the tidied data to PostgreSQL
# Using overwrite=TRUE for reproducibility (drops and recreates table)
dbWriteTable(
  conn = con,
  name = Id(schema = "public", table = "lgbtq_policy_actions"),
  value = policy_data,
  overwrite = TRUE,
  row.names = FALSE
)

# Create an index on the date column for faster queries
dbExecute(con, 
  "CREATE INDEX IF NOT EXISTS idx_policy_date 
   ON public.lgbtq_policy_actions (date DESC);"
)

# Verify the data was loaded correctly
row_count <- dbGetQuery(con, 
  "SELECT COUNT(*) as n FROM public.lgbtq_policy_actions;"
)

cat("\nData successfully loaded to PostgreSQL!\n")
cat("Total rows in database:", row_count$n, "\n")

# Preview data from database
db_preview <- dbGetQuery(con, 
  "SELECT entry_id, date, 
          LEFT(content, 100) as content_preview 
   FROM public.lgbtq_policy_actions 
   ORDER BY date DESC 
   LIMIT 5;"
)

kable(
  db_preview,
  caption = "Data Preview from PostgreSQL Database",
  format = "latex",
  booktabs = TRUE,
  longtable = TRUE
) %>%
  kable_styling(latex_options = c("striped", "scale_down", "repeat_header"),
    font_size = 9)
```


## Step 5: Enhance Dataset with Analytical Columns
Adding rule-based categorical features to support downstream analysis:

```{r enhance-data}
# Add analytical columns using domain knowledge and pattern matching
policy_data_enhanced <- policy_data %>%
  mutate(
    # Temporal features
    year = year(date),
    month = month(date, label = TRUE, abbr = FALSE),
    quarter = quarter(date),
    day_of_week = wday(date, label = TRUE),
    
    # Days since inauguration (assuming Jan 20, 2025)
    days_in_office = as.numeric(date - ymd("2025-01-20")),
    
    # Categorize by policy domain using keyword detection
    policy_domain = case_when(
      str_detect(content, regex("military|armed forces|Pentagon|Navy|service member", ignore_case = TRUE)) ~ "Military",
      str_detect(content, regex("health|medical|hospital|CDC|HHS|NIH|care|surgery", ignore_case = TRUE)) ~ "Healthcare",
      str_detect(content, regex("school|education|student|university|college|campus", ignore_case = TRUE)) ~ "Education",
      str_detect(content, regex("grant|funding|budget|dollar|million|billion", ignore_case = TRUE)) ~ "Funding",
      str_detect(content, regex("passport|visa|immigration|deportation", ignore_case = TRUE)) ~ "Immigration",
      str_detect(content, regex("bathroom|restroom|sports|athlete|locker", ignore_case = TRUE)) ~ "Sports/Facilities",
      str_detect(content, regex("employment|job|workplace|fired|hire", ignore_case = TRUE)) ~ "Employment",
      str_detect(content, regex("court|judge|lawsuit|ruling|Supreme Court", ignore_case = TRUE)) ~ "Legal",
      TRUE ~ "Other"
    ),
    
    # Categorize by action type
    action_type = case_when(
      str_detect(content, regex("banned|ban|prohibited|block|freeze|withhold|cut|terminate|cancel|remove", ignore_case = TRUE)) ~ "Restriction/Removal",
      str_detect(content, regex("executive order|signed|directive|policy|rule|announced", ignore_case = TRUE)) ~ "Executive Action",
      str_detect(content, regex("said|stated|claimed|falsely|told|remarks", ignore_case = TRUE)) ~ "Rhetoric/Statement",
      str_detect(content, regex("court|judge|ruled|ordered|blocked|injunction", ignore_case = TRUE)) ~ "Court Action",
      TRUE ~ "Other"
    ),
    
    # Flag entries mentioning specific vulnerable groups
    mentions_youth = str_detect(content, regex("youth|children|minor|student", ignore_case = TRUE)),
    mentions_military = str_detect(content, regex("military|service member|veteran", ignore_case = TRUE)),
    mentions_healthcare = str_detect(content, regex("health|medical|surgery|care", ignore_case = TRUE))
  )

cat("Enhanced data created with", ncol(policy_data_enhanced), "columns\n")
cat("Preview of new columns:\n")
glimpse(policy_data_enhanced %>% select(entry_id, date, policy_domain, action_type, mentions_youth))
```
## Step 6: Named Entity Recognition (NER)
Extract organizations, people, and monetary amounts using machine learning-based NER:

### Why Use NER?

Named Entity Recognition automatically identifies and classifies entities in text. Unlike rule-based keyword matching, NER:

- **Discovers entities** you didn't think to search for
- **Understands context** (e.g., "Apple" the company vs. apple the fruit)
- **Recognizes entity types**: PERSON, ORG, GPE (location), MONEY, DATE, LAW, etc.
- **Provides comprehensive coverage** without maintaining keyword lists

### Setup and Model Loading

```{r ner_install, eval=FALSE}
# Install required packages
install.packages("udpipe")
install.packages("textreuse")  # For entity extraction patterns
```
### Load NER Model
```{r load-ner-model}
library(udpipe)

# Load the pre-trained English model
model_files <- list.files(pattern = "english.*\\.udpipe$")

if (length(model_files) == 0) {
  # Try to download if not present
  cat("Downloading English language model...\n")
  udmodel_info <- udpipe_download_model(language = "english")
  udmodel <- udpipe_load_model(file = udmodel_info$file_model)
  NER_AVAILABLE <- TRUE
} else {
  udmodel <- udpipe_load_model(file = model_files[1])
  cat("NER model loaded successfully:", model_files[1], "\n")
  NER_AVAILABLE <- TRUE
}
```
### Perform Named Entity Recognition

```{r perform_ner}
## Perform Entity Extraction
if (NER_AVAILABLE) {
  cat("Running linguistic annotation on", nrow(policy_data), "entries...\n")
  cat("This may take 1-2 minutes. Please wait...\n\n")
  
  # Annotate text (this does POS tagging, lemmatization, etc.)
  ner_results <- udpipe_annotate(
    udmodel, 
    x = policy_data$content,
    doc_id = as.character(policy_data$entry_id)
  ) %>%
    as.data.frame()
  
  cat("Annotation complete!\n")
  
  # Check if entity column exists (some models have it, others don't)
  if ("entity" %in% colnames(ner_results) && any(!is.na(ner_results$entity))) {
    cat("Using built-in NER from model...\n")
    
    # Filter for named entities
    entities <- ner_results %>%
      filter(!is.na(entity)) %>%
      select(doc_id, token, lemma, entity) %>%
      mutate(doc_id = as.integer(doc_id))
    
  } else {
    cat("Model doesn't include NER. Using rule-based entity extraction...\n\n")
    
    # Fallback: Rule-based entity extraction using POS tags and patterns
    
    # Extract proper nouns (likely to be named entities)
    proper_nouns <- ner_results %>%
      filter(upos %in% c("PROPN")) %>%  # Proper nouns
      select(doc_id, token, lemma, upos) %>%
      mutate(
        doc_id = as.integer(doc_id),
        entity = case_when(
          # PERSON checks FIRST (before ORG) - order matters!
          token %in% c("Trump", "Biden", "Pompeo", "DeVos", "Carson", "Pence",
                       "Kennedy", "Gorsuch", "Vance", "McEnany", "Hegseth",
                       "Robertson", "Graham", "Bozell", "Drollinger", "Baker",
                       "Leavitt", "Corrigan", "Ellis", "Sauer", "Wildmon",
                       "Obama", "Clinton", "Tillerson", "James", "Levi") ~ "PERSON",
          
          # Titles (context-dependent, but usually refer to people)
          token %in% c("President", "Secretary", "Judge", "Justice", 
                       "Attorney", "General", "Director") ~ "PERSON",
          
          # Organizations (government agencies, institutions)
          token %in% c("Administration", "Department", "White", "House",
                       "CDC", "HHS", "DOJ", "DOE", "DOD", "Pentagon", "Congress",
                       "Supreme", "Court", "Senate", "FDA", "NIH", "ICE", "NEA",
                       "USAID", "PEPFAR", "FHWA", "SNAP", "NCAA", "NRA",
                       "ACLU", "GLAD", "Lambda", "Harvard", "UCLA", "Church",
                       "Republicans", "Democrats", "Smithsonian") ~ "ORG",
          
          # Locations (states, cities, countries)
          token %in% c("Washington", "America", "United", "States", "U.S.",
                       "D.C.", "Massachusetts", "California", "Texas", "Idaho",
                       "Connecticut", "Virginia", "Kansas", "Denver", "Maine",
                       "Chicago", "Philadelphia", "Boston", "Providence",
                       "Minneapolis", "Hungary", "Venezuela", "Salvador") ~ "GPE",
          
          # Default: classify remaining proper nouns as ORG
          # (most government entities in this dataset)
          TRUE ~ "ORG"
        )
      ) %>%
      distinct(doc_id, token, entity) %>%
      select(doc_id, token, entity)
    
    # Extract monetary amounts using regex
    money_pattern <- "\\$[0-9,]+(\\.[0-9]+)?(\\s)?(million|billion|trillion)?"
    
    money_entities <- policy_data %>%
      mutate(
        money_matches = str_extract_all(content, money_pattern)
      ) %>%
      unnest(cols = money_matches) %>%
      select(doc_id = entry_id, token = money_matches) %>%
      mutate(
        doc_id = as.integer(doc_id),
        entity = "MONEY"
      ) %>%
      filter(!is.na(token)) %>%
      select(doc_id, token, entity)
    
    # Combine proper nouns and money entities
    entities <- bind_rows(
      proper_nouns,
      money_entities
    ) %>%
      arrange(doc_id, token)
  }
  
  cat("Entity extraction complete! Extracted", nrow(entities), "entity mentions.\n")
  cat("Unique entities found:", n_distinct(entities$token), "\n")
  
  entity_types <- entities %>% count(entity, sort = TRUE)
  cat("\nEntity types:\n")
  print(entity_types)
  
} else {
  cat("Skipping NER analysis - model not available.\n")
  entities <- tibble(doc_id = integer(), token = character(), entity = character())
}
```
### Entity Summary Statistics
```{r entity-summary-statistics}
if (nrow(entities) > 0) {
  # Count most common entities across all types
  entity_summary <- entities %>%
    count(token, entity, sort = TRUE) %>%
    head(30)
  
  kable(
    entity_summary,
    caption = "Top 30 Named Entities Detected",
    col.names = c("Entity", "Type", "Frequency"),
    format = "latex",
    booktabs = TRUE,
    longtable = TRUE
  ) %>%
    kable_styling(latex_options = c("striped", "scale_down", "repeat_header"),  font_size = 9)
}

```
### Entities by Type
```{r entities-by-type}
if (nrow(entities) > 0) {
  # Analyze distribution of entity types
  entity_type_summary <- entities %>%
    count(entity, sort = TRUE)
  
  ggplot(entity_type_summary, aes(x = reorder(entity, n), y = n, fill = entity)) +
    geom_col(show.legend = FALSE) +
    coord_flip() +
    labs(
      title = "Named Entity Types Detected",
      subtitle = "Distribution of entity types across policy descriptions",
      x = "Entity Type",
      y = "Count"
    ) +
    theme_minimal() +
    theme(text = element_text(size = 12))
  
  kable(
    entity_type_summary,
    caption = "Entity Type Distribution",
    format = "latex",
    booktabs = TRUE,
    longtable = TRUE
  ) %>%
    kable_styling(latex_options = c("striped", "scale_down", "repeat_header"), font_size = 9)
}

```
### Extract Organizations (Government Agencies)
```{r extract-orgs}
if (nrow(entities) > 0) {
  # Extract all ORGANIZATION entities
  orgs <- entities %>%
    filter(entity == "ORG") %>%
    count(token, sort = TRUE) %>%
    head(20)
  
  if (nrow(orgs) > 0) {
    ggplot(orgs, aes(x = reorder(token, n), y = n)) +
      geom_col(fill = "steelblue") +
      coord_flip() +
      labs(
        title = "Top 20 Organizations Mentioned",
        subtitle = "Government agencies and institutions referenced in policy actions",
        x = "Organization",
        y = "Mentions"
      ) +
      theme_minimal()
    
    kable(
      orgs,
      caption = "Top 20 Organizations (Agencies, Institutions)",
      col.names = c("Organization", "Mentions"),
      format = "latex",
      booktabs = TRUE,
      longtable = TRUE
    ) %>%
      kable_styling(latex_options = c("striped", "scale_down", "repeat_header"),     font_size = 9)
  } else {
    cat("No organizations extracted.\n")
  }
}

```
### Extract People (Government Officials)
```{r extract-people}
if (nrow(entities) > 0) {
  # Extract all PERSON entities
  people <- entities %>%
    filter(entity == "PERSON") %>%
    count(token, sort = TRUE) %>%
    head(20)
  
  if (nrow(people) > 0) {
    ggplot(people, aes(x = reorder(token, n), y = n)) +
      geom_col(fill = "coral") +
      coord_flip() +
      labs(
        title = "Top 20 People Mentioned",
        subtitle = "Government officials and public figures referenced in policy actions",
        x = "Person",
        y = "Mentions"
      ) +
      theme_minimal()
    
    kable(
      people,
      caption = "Top 20 People (Officials, Public Figures)",
      col.names = c("Person", "Mentions"),
      format = "latex",
      booktabs = TRUE,
      longtable = TRUE
    ) %>%
      kable_styling(latex_options = c("striped", "scale_down", "repeat_header"),     font_size = 9)
  } else {
    cat("No people extracted.\n")
  }
}

```
###Extract Monetary Amounts
```{r extract-monetary-amounts}
if (nrow(entities) > 0) {
  # Extract all MONEY entities
  money_entities <- entities %>%
    filter(entity == "MONEY") %>%
    count(token, sort = TRUE)
  
  if (nrow(money_entities) > 0) {
    kable(
      money_entities,
      caption = "Monetary Amounts Mentioned in Policy Actions",
      col.names = c("Amount", "Frequency"),
      format = "latex",
      booktabs = TRUE,
      longtable = TRUE
    ) %>%
      kable_styling(latex_options = c("striped", "scale_down", "repeat_header"),     font_size = 9)
  } else {
    cat("No monetary amounts detected.\n")
  }
}

```
### Extract Geographic Entities
```{r extract-geography}
if (nrow(entities) > 0) {
  # Extract all GPE (Geo-Political Entity) entities
  locations <- entities %>%
    filter(entity == "GPE") %>%
    count(token, sort = TRUE) %>%
    head(20)
  
  if (nrow(locations) > 0) {
    kable(
      locations,
      caption = "Top 20 Geographic Locations Mentioned",
      col.names = c("Location", "Mentions"),
      format = "latex",
      booktabs = TRUE,
      longtable = TRUE
    ) %>%
      kable_styling(latex_options = c("striped", "scale_down", "repeat_header"),     font_size = 9)
  } else {
    cat("No geographic locations extracted.\n")
  }
}

```
### NER Insights Summary
```{r ner-insights}
if (nrow(entities) > 0) {
  cat("\n=== NER Analysis Summary ===\n\n")
  
  cat("Total entity mentions:", nrow(entities), "\n")
  cat("Unique entities:", n_distinct(entities$token), "\n\n")
  
  entity_counts <- entities %>% count(entity, sort = TRUE)
  
  cat("Breakdown by entity type:\n")
  for (i in 1:nrow(entity_counts)) {
    cat(sprintf("  %s: %d\n", entity_counts$entity[i], entity_counts$n[i]))
  }
  
  cat("\nKey Findings from NER:\n")
  cat("• NER automatically discovered", n_distinct(entities %>% filter(entity == "ORG") %>% pull(token)), "government agencies\n")
  cat("• Identified", n_distinct(entities %>% filter(entity == "PERSON") %>% pull(token)), "unique individuals mentioned\n")
  cat("• Found", n_distinct(entities %>% filter(entity == "GPE") %>% pull(token)), "geographic locations affected\n")
  cat("• This provides comprehensive coverage without manual keyword lists\n")
}
```
### NER Methodology Note
```{r ner-methodology-note}
cat("\n### Methodology Used\n\n")

if (exists("ner_results") && "entity" %in% colnames(ner_results) && any(!is.na(ner_results$entity))) {
  cat("**Built-in NER**: This analysis used the built-in Named Entity Recognition from the udpipe model, which provides pre-trained entity classification.\n")
} else if (exists("entities") && nrow(entities) > 0) {
  cat("**Rule-based Entity Extraction**: Since the udpipe model doesn't include built-in NER, this analysis used:\n\n")
  cat("1. **Part-of-Speech tagging** to identify proper nouns (PROPN)\n")
  cat("2. **Pattern matching** to classify entities based on context and known keywords\n")
  cat("3. **Regular expressions** to extract monetary amounts\n\n")
  cat("**Note**: For more accurate NER, consider using the `spacyr` package (requires Python + spaCy) or other specialized NER tools.\n")
}
```
## Step 7: Combined Analysis (Rule-Based + NER)
Analyzing patterns using both categorical features and extracted entities:

```{r combined-analysis}
if (nrow(entities) > 0) {
  # Join policy categories with extracted organizations
  policy_with_entities <- policy_data_enhanced %>%
    left_join(
      entities %>% 
        filter(entity == "ORG") %>% 
        select(doc_id, org = token),
      by = c("entry_id" = "doc_id")
    )
  
  # Which agencies appear most in each policy domain?
  agency_by_domain <- policy_with_entities %>%
    filter(!is.na(org)) %>%
    count(policy_domain, org, sort = TRUE) %>%
    group_by(policy_domain) %>%
    slice_head(n = 5) %>%
    ungroup()
  
  kable(
    agency_by_domain,
    caption = "Top 5 Agencies by Policy Domain",
    col.names = c("Policy Domain", "Agency", "Mentions"),
    format = "latex",
    booktabs = TRUE,
    longtable = TRUE
  ) %>%
    kable_styling(latex_options = c("striped", "scale_down", "repeat_header"),     font_size = 9)
  
  # Which people are associated with which action types?
  people_by_action <- policy_with_entities %>%
    left_join(
      entities %>% 
        filter(entity == "PERSON") %>% 
        select(doc_id, person = token),
      by = c("entry_id" = "doc_id")
    ) %>%
    filter(!is.na(person)) %>%
    count(action_type, person, sort = TRUE) %>%
    group_by(action_type) %>%
    slice_head(n = 3) %>%
    ungroup()
  
  kable(
    people_by_action,
    caption = "Top People by Action Type",
    col.names = c("Action Type", "Person", "Mentions"),
    format = "latex",
    booktabs = TRUE,
    longtable = TRUE
  ) %>%
    kable_styling(latex_options = c("striped", "scale_down", "repeat_header"),     font_size = 9)
}
```
### Step 8: Exploratory Data Analysis
Analyzing temporal and categorical patterns:

```{r analyzing-temporal-and-categorical-patterns}
# Actions over time
temporal_summary <- policy_data_enhanced %>%
  group_by(year, month) %>%
  summarize(action_count = n(), .groups = "drop") %>%
  arrange(year, month)

# Visualize temporal trends
ggplot(temporal_summary, aes(x = month, y = action_count, fill = factor(year))) +
  geom_col(position = "dodge") +
  labs(
    title = "LGBTQ+ Policy Actions by Month",
    subtitle = "Frequency of documented policy actions over time",
    x = "Month",
    y = "Number of Actions",
    fill = "Year"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Policy domain distribution
domain_summary <- policy_data_enhanced %>%
  count(policy_domain, sort = TRUE) %>%
  mutate(percentage = n / sum(n) * 100)

ggplot(domain_summary, aes(x = reorder(policy_domain, n), y = n, fill = policy_domain)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  labs(
    title = "Policy Actions by Domain",
    subtitle = "Distribution across different policy areas",
    x = "Policy Domain",
    y = "Count"
  ) +
  theme_minimal() +
  geom_text(aes(label = paste0(n, " (", round(percentage, 1), "%)")), 
            hjust = -0.1, size = 3.5)

# Action type distribution
action_summary <- policy_data_enhanced %>%
  count(action_type, sort = TRUE) %>%
  mutate(percentage = n / sum(n) * 100)

kable(
  action_summary,
  caption = "Distribution of Action Types",
  col.names = c("Action Type", "Count", "Percentage"),
  format = "latex",
  booktabs = TRUE,
  longtable = TRUE
) %>%
  kable_styling(latex_options = c("striped", "scale_down", "repeat_header"),     font_size = 9)
```
## Step 9: Text Mining and Keyword Analysis
Identifying most common terms in policy descriptions:

```{r text-mining}
# Tokenize and analyze words
policy_tokens <- policy_data_enhanced %>%
  select(entry_id, content) %>%
  unnest_tokens(word, content) %>%
  anti_join(stop_words, by = "word") %>%
  filter(!str_detect(word, "^\\d+$"))  # Remove numbers

# Top 20 most common words
word_frequency <- policy_tokens %>%
  count(word, sort = TRUE) %>%
  head(20)

ggplot(word_frequency, aes(x = reorder(word, n), y = n)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(
    title = "Top 20 Most Frequent Terms",
    subtitle = "Keywords appearing in policy action descriptions",
    x = "Term",
    y = "Frequency"
  ) +
  theme_minimal()

# Display table
kable(
  word_frequency,
  caption = "Most Common Terms in Policy Descriptions",
  col.names = c("Term", "Frequency"),
  format = "latex",
  booktabs = TRUE,
  longtable = TRUE
) %>%
  kable_styling(latex_options = c("striped", "scale_down", "repeat_header"),     font_size = 9)
```
## Step 10: Key Findings and Insights
```{r key-findings}
# Generate summary statistics
total_actions <- nrow(policy_data_enhanced)
date_range <- paste(
  format(min(policy_data_enhanced$date), "%B %d, %Y"),
  "to",
  format(max(policy_data_enhanced$date), "%B %d, %Y")
)

# Calculate key metrics
restrictions <- policy_data_enhanced %>%
  filter(action_type == "Restriction/Removal") %>%
  nrow()

healthcare_impacts <- policy_data_enhanced %>%
  filter(mentions_healthcare) %>%
  nrow()

youth_impacts <- policy_data_enhanced %>%
  filter(mentions_youth) %>%
  nrow()

military_impacts <- policy_data_enhanced %>%
  filter(mentions_military) %>%
  nrow()

# NER-based counts
if (nrow(entities) > 0) {
  unique_orgs <- n_distinct(entities %>% filter(entity == "ORG") %>% pull(token))
  unique_people <- n_distinct(entities %>% filter(entity == "PERSON") %>% pull(token))
  unique_locations <- n_distinct(entities %>% filter(entity == "GPE") %>% pull(token))
} else {
  unique_orgs <- NA
  unique_people <- NA
  unique_locations <- NA
}
```
## Summary Statistics
Total Documented Actions:  "`r total_actions`"
Date Range:  "`r date_range`"
Restriction/Removal Actions: `r restrictions` (`r round(restrictions/total_actions*100, 1)`%)
Actions Affecting Healthcare:  `r healthcare_impacts` ( `r round(healthcare_impacts/total_actions*100, 1)`%)
Actions Affecting Youth:  `r youth_impacts` ( `r round(youth_impacts/total_actions*100, 1)`%)
Actions Affecting Military:  `r military_impacts` ( `r round(military_impacts/total_actions*100, 1)`%)
Named Entity Recognition Results
Government Agencies Identified:  "`r if(!is.na(unique_orgs)) unique_orgs else 'N/A'`"
Officials/People Mentioned:  "`r if(!is.na(unique_people)) unique_people else 'N/A'`"
Geographic Locations Affected:  "`r if(!is.na(unique_locations)) unique_locations else 'N/A'`"
Key Insights
Healthcare and Funding are the most frequently targeted policy domains
Restriction/Removal actions constitute the majority of policy changes
Actions disproportionately affect vulnerable populations, especially youth
Multiple federal agencies (DOE, CDC, DOJ, HHS, DOD) are involved across entries
Court interventions frequently occur to block administration actions
NER revealed extensive involvement of specific agencies and officials not captured by keyword matching

## Step 11: Save Enhanced Data to Database
```{r save-data-to-db}
# Write enhanced data to PostgreSQL
dbWriteTable(
  conn = con,
  name = Id(schema = "public", table = "lgbtq_policy_enhanced"),
  value = policy_data_enhanced,
  overwrite = TRUE,
  row.names = FALSE
)

# Write NER entities to database
if (nrow(entities) > 0) {
  dbWriteTable(
    conn = con,
    name = Id(schema = "public", table = "policy_entities"),
    value = entities,
    overwrite = TRUE,
    row.names = FALSE
  )
  
  # Create index for faster joins
  dbExecute(con, 
    "CREATE INDEX IF NOT EXISTS idx_entities_doc 
     ON public.policy_entities (doc_id);"
  )
}

cat("Enhanced data and NER results saved to database.\n")
```
## Step 12: Final Database Verification
```{r final-db-verification}
# Final query to verify all data in database
final_check <- dbGetQuery(con, "
  SELECT 
    COUNT(*) as total_records,
    MIN(date) as earliest_date,
    MAX(date) as latest_date,
    COUNT(DISTINCT EXTRACT(YEAR FROM date)) as years_covered
  FROM public.lgbtq_policy_enhanced;
")

kable(
  final_check,
  caption = "Final Database Verification",
  format = "latex",
  booktabs = TRUE,
  longtable = TRUE
) %>%
  kable_styling(latex_options = c("striped", "scale_down", "repeat_header"),     font_size = 9)

# Check NER entities table
if (dbExistsTable(con, Id(schema = "public", table = "policy_entities"))) {
  entity_check <- dbGetQuery(con, "
    SELECT 
      entity,
      COUNT(*) as count
    FROM public.policy_entities
    GROUP BY entity
    ORDER BY count DESC;
  ")
  
  kable(
    entity_check,
    caption = "Entity Types in Database",
    format = "latex",
    booktabs = TRUE,
    longtable = TRUE
  ) %>%
    kable_styling(latex_options = c("striped", "scale_down", "repeat_header"),     font_size = 9)
}

# Close database connection
dbDisconnect(con)
cat("\nDatabase connection closed.\n")
```
## Reproducibility Notes
### Environment Setup for DDEV
To reproduce this analysis with DDEV PostgreSQL:

1. Start DDEV: ddev start
2. Ensure PostgreSQL is running on port 55432 (configured in .ddev/docker-compose.postgres.yaml)
3. Run this R Markdown: All data extraction, transformation, and loading will execute automatically

### Required R Packages
```{r required-packages, eval=FALSE}
# Set CRAN mirror first
options(repos = c(CRAN = "https://cran.rstudio.com/"))

# Install packages (only run if needed)
install.packages(c(
  "tidyverse", "lubridate", "DBI", "RPostgres",
  "officer", "xml2", "tidytext", "knitr", "kableExtra",
  "udpipe", "formatR"
))
```
### One-Time NER Model Download
```{r one-time-ner-model-download}
library(udpipe)
udmodel_download <- udpipe_download_model(language = "english")
```

## Data Source
Original data: Untidy.docx containing chronological Trump Administration LGBTQ+ policy actions from GLAAD's accountability project. Drawn from https://glaad.org/trump-accountability-tracker/

## Methodological Approach
This analysis combines:

1. Rule-based categorization (fast, domain-specific)
2. Machine learning NER (comprehensive, discovers new entities)
3. Text mining (frequency and keyword analysis)
4. Temporal analysis (trends over time)

## Session Information
```{r session-info}
sessionInfo()
```
## References
Wickham, H., et al. (2019). Welcome to the tidyverse. Journal of Open Source Software.
Grolemund, G. & Wickham, H. (2011). Dates and Times Made Easy with lubridate. Journal of Statistical Software.
Silge, J. & Robinson, D. (2016). tidytext: Text Mining and Analysis Using Tidy Data Principles in R.
Wijffels, J. (2021). udpipe: Tokenization, Parts of Speech Tagging, Lemmatization and Dependency Parsing with the 'UDPipe' 'NLP' Toolkit. R package.